SNP Analysis

### 2020-07-14 snp/SNP_Prediction_Analysis.Rmd
Updates to the paper, and generating a standardized patient/sample database for metadata

p = ggplot(qcdata.endo, aes(Purity_mean, ASCAT.SCA.ratio_mean)) + facet_grid(~SampleType) +
  geom_point(aes(color=ASCAT.SCA.ratio_mean < 0.01 & Purity_mean > 0.95 ), show.legend = F) + 
  labs(title='Merged biopsies per-timepoint', y='mean(ASCAT.SCA.ratio)', x='mean(ASCAT purity)')
p
ggsave('plots/snp/sca_purity_per_tp.png', p, width=4, height=4, units='in')

@@

Despite that, we still see a fairly consistent high/low risk prediction for NP vs P.  However, HGD in NP patients predicts at a rate consistent with progression.
```{r echo=F}
preds = preds %>% mutate(Pathology = factor(Pathology, levels=c('NDBE','ID','LGD','HGD','EAC'), ordered = T))
rp = plot.risk.by.path(preds) + labs(y='% Classified', title='Samples classified by pathology')
rp

@@

rocBE = pROC::roc(Status ~ Prob, data=preds, auc=T, ci=T, of='thresholds')
roc.plot(rocBE,title='SNP ROC, BE biopsies',incS = T, 0.3)
rocBE$model = 'per Biopsy'

@@

rocEX = pROC::roc( Path.Status ~ Prob, data=pred.qc, auc=T, ci=T, of='thresholds')
roc.plot(rocEX) + labs(title='BE biopsies: Path status (HGD/IMC)')
ggsave(filename='plots/snp/path_status_roc.png', plot = roc.plot(rocEX) + labs(title='BE biopsies', subtitle='Path status (HGD/IMC)') + theme(text=element_text(size=12)), width=4, height=4, units='in', dpi=300)
rocEX$model = 'HGD/IMC progressed'

@@

# Exclude the low SCA patients
rocpp = pROC::roc(Status ~ Prob, data=filter(pred.qc, !lowsca), auc=T, ci=T, of='thresholds')
roc.plot(rocpp, 'BE biopsies: excl. low SCA samples')
ggsave(filename='plots/snp/auc_ex_lowsca_roc.png', plot = roc.plot(rocpp, 'BE biopsies: excl. low SCA samples') + theme(text=element_text(size=12)), width=5, height=5, units='in', dpi=300)
rocpp$model = 'All excl. low SCA'

# Exclude all of the above
rocexall = pROC::roc(Status ~ Prob, data=filter(pred.qc, !lowsca & !wgd.NP), auc=T, ci=T, of='thresholds')
roc.plot(rocexall, 'BE biopsies: excl. low SCA & WGD NP')
ggsave(filename='plots/snp/auc_ex_lowsca_wgs_roc.png', plot =roc.plot(rocexall, 'BE biopsies: excl. low SCA & WGD NP') + theme(text=element_text(size=12)), width=4, height=4, units='in', dpi=300)
rocexall$model = 'excl. low SCA & (NP) WGD'

# Timepoint 1 only
roct1 = pROC::roc(Status ~ Prob, data=filter(pred.qc, Timepoint == 'T1' & !lowsca & !wgd.NP), auc=T, ci=T, of='thresholds')
roc.plot(roct1, 'BE biopsies: Timepoint T1')
ggsave(filename='plots/snp/auc_t1_roc.png', plot=roc.plot(roct1, 'BE biopsies: Timepoint T1') + theme(text=element_text(size=12)), width=5, height=5, units='in', dpi=300)
roct1$model = 'Timepoint 1 (with exclusions)'

# Timepoint 2 only
roct2 = pROC::roc(Status ~ Prob, data=subset(pred.qc, Timepoint != 'T1' & !lowsca & !wgd.NP), auc=T, ci=T, of='thresholds')
roc.plot(roct2, 'BE biopsies: Timepoint T2')
ggsave(filename='plots/snp/auc_t2_roc.png', plot=roc.plot(roct2, 'BE biopsies: Timepoint T2') + theme(text=element_text(size=12)), width=5, height=5, units='in', dpi=300)
roct2$model = 'Timepoint 2  (with exclusions)'

m = bind_rows(
      get.roc.stat(rocBE,0.3),
      get.roc.stat(rocEX,0.3),
      get.roc.stat(rocexall,0.3),
      get.roc.stat(rocpp,0.3),
      get.roc.stat(roct1,0.3),
      get.roc.stat(roct2,0.3)
) %>% dplyr::mutate_if(is.numeric, list(~round(.,2))) %>%
  dplyr::mutate(model = factor(model, levels = c(rocBE$model, rocEX$model, rocexall$model, rocpp$model, roct1$model, roct2$model), ordered = T))

@@

```{r regression, echo=F, message=F, eval=F}
# Regression using SNP Data
folds = 10; splits = 5
## ----- All ----- ##
cache.dir = '~/Data/BarrettsProgressionRisk//Analysis/SNP/no-low/'
file = paste(cache.dir, 'all.pt.alpha.Rdata', sep='/')
if (file.exists(file)) {
 message(paste("loading", file))
 load(file, verbose=T)
 model.performance(performance.at.1se, coefs, folds, splits)$plot + labs(title='SNP CV models')
} else {
  warning("Run 'snp-regression.R' first")
}
#save(arrayDf,labels,patient.info, file=paste(cache.dir,'snp-models.Rdata',sep='/'))

>>>>

plotL <- c('BE' = paste0('BE\n(n=',nrow(filter(qcdata.endo, SampleType == 'BE')),')' ),
           'Blood/Gastric Normal' = paste0('Blood/Gastric Normal\n(n=', nrow(filter(qcdata.endo, SampleType == 'Blood/Gastric Normal')), ')' ))

qcdata.endo %>% ungroup %>% dplyr::select(Status, Purity_mean, ASCAT.SCA.ratio_mean, SampleType) %>% 
  dplyr::mutate(Status = recode(Status, 'NP' = 'Non-Progressor', 'P'='Progressor')) %>% dplyr::arrange(Status, SampleType) %>% write_tsv('plots/exfig3d.tsv')

p = ggplot(qcdata.endo, aes(Purity_mean, ASCAT.SCA.ratio_mean)) + facet_grid(~SampleType, labeller = labeller(SampleType = plotL)) +
  geom_point(aes(color=ASCAT.SCA.ratio_mean < 0.01 & Purity_mean > 0.95 ), show.legend = F) + 
  labs(title='Merged biopsies per-timepoint', y='mean(ASCAT.SCA.ratio)', x='mean(ASCAT purity)')
p
ggsave('plots/snp/sca_purity_per_tp.png', p, width=4, height=4, units='in')

@@

Despite that, we still see a fairly consistent high/low risk prediction for NP vs P.  However, HGD in NP patients predicts at a rate consistent with progression.
```{r echo=F}
preds = preds %>% mutate(Pathology = factor(Pathology, levels=c('NDBE','ID','LGD','HGD','EAC'), ordered = T))

preds %>% dplyr::select(Prob, RR, Risk, Pathology, Status) %>% dplyr::rename('Probability' = 'Prob', 'Relative Risk' = 'RR') %>%
  dplyr::mutate(Status = recode(Status, 'NP'='Non-Progressor', 'P' = 'Progressor') ) %>% dplyr::arrange(Status, Pathology) %>%
  write_tsv('plots/exfig3b.tsv')

rp = plot.risk.by.path(preds) + labs(y='% Classified', title='Samples classified by pathology')
rp

@@

rocBE = pROC::roc(Status ~ Prob, data=preds, auc=T, ci=T, of='thresholds')
roc.plot(rocBE,title='SNP ROC, BE biopsies',incS = T, 0.3)
rocBE$model = 'per Biopsy'
rocBE$n = nrow(preds)

preds %>% dplyr::select(Status, Prob, RR) %>% dplyr::rename('Probability' = 'Prob', 'Relative Risk' = 'RR') %>%
  dplyr::mutate(Status = recode(Status, 'NP'='Non-Progressor', 'P' = 'Progressor') ) %>% dplyr::arrange(Status, Probability) %>%
  write_tsv('plots/exfig3a.tsv')

@@

rocEX = pROC::roc( Path.Status ~ Prob, data=pred.qc, auc=T, ci=T, of='thresholds')
roc.plot(rocEX) + labs(title='BE biopsies: Path status (HGD/IMC)')
ggsave(filename='plots/snp/path_status_roc.png', plot = roc.plot(rocEX) + labs(title='BE biopsies', subtitle='Path status (HGD/IMC)') + theme(text=element_text(size=12)), width=4, height=4, units='in', dpi=300)
rocEX$model = 'HGD/IMC progressed'
rocEX$n = nrow(pred.qc)

@@

# Exclude the low SCA patients
rocpp = pROC::roc(Status ~ Prob, data=filter(pred.qc, !lowsca), auc=T, ci=T, of='thresholds')
roc.plot(rocpp, 'BE biopsies: excl. low SCA samples')
ggsave(filename='plots/snp/auc_ex_lowsca_roc.png', plot = roc.plot(rocpp, 'BE biopsies: excl. low SCA samples') + theme(text=element_text(size=12)), width=5, height=5, units='in', dpi=300)
rocpp$model = 'All excl. low SCA'
rocpp$n = nrow(filter(pred.qc, !lowsca))

# Exclude all of the above
rocexall = pROC::roc(Status ~ Prob, data=filter(pred.qc, !lowsca & !wgd.NP), auc=T, ci=T, of='thresholds')
roc.plot(rocexall, 'BE biopsies: excl. low SCA & WGD NP')
ggsave(filename='plots/snp/auc_ex_lowsca_wgs_roc.png', plot =roc.plot(rocexall, 'BE biopsies: excl. low SCA & WGD NP') + theme(text=element_text(size=12)), width=4, height=4, units='in', dpi=300)
rocexall$model = 'excl. low SCA & (NP) WGD'
rocexall$n = nrow(filter(pred.qc, !lowsca & !wgd.NP))

# Timepoint 1 only
roct1 = pROC::roc(Status ~ Prob, data=filter(pred.qc, Timepoint == 'T1' & !lowsca & !wgd.NP), auc=T, ci=T, of='thresholds')
roc.plot(roct1, 'BE biopsies: Timepoint T1')
ggsave(filename='plots/snp/auc_t1_roc.png', plot=roc.plot(roct1, 'BE biopsies: Timepoint T1') + theme(text=element_text(size=12)), width=5, height=5, units='in', dpi=300)
roct1$model = 'Timepoint 1 (with exclusions)'
roct1$n = nrow(filter(pred.qc, Timepoint == 'T1' & !lowsca & !wgd.NP))

# Timepoint 2 only
roct2 = pROC::roc(Status ~ Prob, data=filter(pred.qc, Timepoint != 'T1' & !lowsca & !wgd.NP), auc=T, ci=T, of='thresholds')
roc.plot(roct2, 'BE biopsies: Timepoint T2')
ggsave(filename='plots/snp/auc_t2_roc.png', plot=roc.plot(roct2, 'BE biopsies: Timepoint T2') + theme(text=element_text(size=12)), width=5, height=5, units='in', dpi=300)
roct2$model = 'Timepoint 2  (with exclusions)'
roct2$n = nrow(filter(pred.qc, Timepoint != 'T1' & !lowsca & !wgd.NP))

m = bind_rows(
      get.roc.stat(rocBE,0.3),
      get.roc.stat(rocEX,0.3),
      get.roc.stat(rocexall,0.3),
      get.roc.stat(rocpp,0.3),
      get.roc.stat(roct1,0.3),
      get.roc.stat(roct2,0.3)
) %>% dplyr::mutate_if(is.numeric, list(~round(.,2))) %>%
  dplyr::mutate(model = factor(model, levels = c(rocBE$model, rocEX$model, rocexall$model, rocpp$model, roct1$model, roct2$model), ordered = T)) %>%
  dplyr::mutate(model = paste0(model,'\n(n=',n,')'))

m %>% dplyr::mutate(model = sub('\n.*', '', model)) %>% write_tsv('plots/exfig3c.tsv')

@@

```{r regression, echo=F, message=F, eval=F}
# Regression using SNP Data
folds = 10; splits = 5
## ----- All ----- ##
cache.dir = '~/Data/BarrettsProgressionRisk//Analysis/SNP/no-low/'
file = paste(cache.dir, 'all.pt.alpha.Rdata', sep='/')
if (file.exists(file)) {
 message(paste("loading", file))
 load(file, verbose=T)

 performance.at.1se %>% bind_rows() %>% dplyr::mutate(penalty = names(performance.at.1se)[-1], 'non-zero coefs'=sapply(coefs, nrow)[-1]) %>% write_tsv('plots/exfig4a.tsv')
 p = model.performance(performance.at.1se, coefs, folds, splits)$plot + labs(title='SNP CV models') + labs(subtitle = paste0('(n=',nrow(df),')'))
 p
 ggsave('plots/snp/model_perf.png', plot=p, width=4, height=4, dpi=300, units='in')
 rm(p)
} else {
  warning("Run 'snp-regression.R' first")
}
#save(arrayDf,labels,patient.info, file=paste(cache.dir,'snp-models.Rdata',sep='/'))
